### 2019-07-06 Memoria/Chapters/NewProposals.tex
Cambios en memoria y en paper DILS

Finding the optimal partition in a dataset, with respect to any kind of reasonable criteria, is known to be a $\mathbf{NP}$-hard problem. As we have already mentioned in Section \ref{sec:FeasibilityProblem}, the incorporation of constraints may modify the complexity of the clustering problem, making it $\mathbf{NP}$-complete and therefore intractable \cite{davidson2005clustering}. Bearing this in mind, in this chapter we propose two new approaches to find approximate solutions to the constrained clustering problem. In Section \ref{sec:CCSHADE} a \acf{DE} based approach is presented, while in Section \ref{sec:DILS} a new \acf{LS} variant is discussed, as well as its application to constrained clustering.

@@

\acs{DE} is an evolution-based algorithm that has proven to be excellent in real-domain problem solving \cite{das2011differential}. In this section we propose the first \acs{DE} application to find quality solutions for the constrained clustering problem. In particular, we will take as basis for our proposal one of the \acs{DE} variants which showed excellent behavior in worldwide competitions, known as SHADE \cite{molina2018insight}. We will make use of the Random-key concept from \acs{BRKGA}, along with proposing a new fitness function, to build the already mentioned \acs{DE} application.

\subsection{A Glimpse to Differential Evolution}

@@

To guide the optimization process, we use a fitness function. The task of \acs{DE} is to seek the parameter vector $p_i^*$ that minimizes a function such that $f(p_i)(f: \Omega \subseteq \mathfrak{R}^{v} \rightarrow \mathfrak{R})$, where $f(p_i^*) < f(p_i)$ for all $p_i \in \Omega$ with $\Omega$ a nonempty finite set that is the search domain.

\begin{figure}[!h]
	\centering
	\includegraphics[scale=0.3]{gfx/NewProp/SHADE/DEop.png}
	\caption[The optimization process of DE.]{The optimization process of \acs{DE} \cite{das2011differential}.}\label{fig:DE}
\end{figure}

@@

Even though \acs{DE} was a major breakthrough in terms of real optimization, many variants have been developed since its invention that provide better results. We will focus on the SHADE variant, which adaptively optimizes some of the parameters of \acs{DE}. To carry out this task SHADE uses a historic record of values for these parameters that allows it to set them in favor of the optimization process, based on how good the algorithm behaved with certain parameter values in the past (see Section \ref{sec:SHADE}).

\subsection{Brief Review of the BRKGA Algorithm} \label{sec:brkga}

@@

In a \acs{GA}, a population $P$ of a given number of solutions \sloppy $p_i = \{p_{[i,1]}, \cdots, p_{[i,v]}\}$ is first initialized; these solutions are also called chromosomes or individuals. Each element of a chromosome is called gene and the value it takes is known as allele. To simulate the natural processes of evolution and natural selection, the set of evolutionary operators that \acsp{GA} make use of is similar to that used by \acs{DE}: crossover, mutation and selection. The crossover operator consists in combining two different chromosomes (parents) from the population and obtaining a new chromosome (offspring) which inherits alleles for its genes from both parents. The mutation operator is used by \acsp{GA} to avoid local optima, and it consists in introducing random changes into the population to favor diversity. The selection operator is used to determine whether a chromosome is suitable to survive for the next generation, and to accomplish this task it employs a fitness function that assigns a value to each chromosome to assess its quality as a solution for the problem. \acsp{GA} apply these operators through a series of generations to hopefully obtain a solution for the problem to solve that is close to the optimal \cite{fausto2019ants}.

\acf{RKGA} emerged in \cite{bean1994genetic} as a solid approach to high constrained problems. It uses vectors of random numbers to represent solutions to the problem (individuals), which are used as keys to obtain the actual solutions. Random-key vectors are sampled from the space $[0,1]^{v}$, which is used as a surrogate space for the literal solution space. Points in the random-key space are transferred to the literal solution space via a decoder for fitness evaluation. A decoder is a deterministic procedure that allows us to map random-key vectors to the literal solution space. We can build a decoder that always provides as a result a feasible solution to the problem we want to solve, thus avoiding the feasibility problem that we found in classic \acsp{GA}.

The \acf{BRKGA} was first proposed in \cite{gonccalves2011biased} as a variant of the \acs{RKGA}. In the \acs{BRKGA} optimization process, first, the population $P$ of $|P|$ vectors of random-keys is initialized; this will be the initial generation. Each random-key vector $p_i$ has $v$ random-keys in it. The population is sorted by the fitness value $f_i$ of each $p_i$ to select the first $P_e$ individuals, this is, the elite of the population. The elite will be preserved in the next generation without modification. To introduce diversity, a number $P_m$ of new random-keys vectors are also included in the next generation. The remaining individuals of the new generation ($|P| - P_e - P_m$) are obtained through crossovers between elite and non-elite parents.

@@

\subsection{BRKGA optimization scheme for Constrained Clustering} \label{sec:AdaptationofBRKGA}

@@

In this scheme each random-key vector has $n$ random-keys in it ($v = n$). The decoder divides the interval $[0,1]$ in $k$ intervals, so there exists a correspondence between each random-key (instance $x_i$) and the integer (label $l_i$) corresponding to the interval which it lies in. Table \ref{tab:decodingrk} shows an example of random-key decoding for a dataset with 10 instances ($n$ = 10) and $k = 3$. Note that extreme values 0 and 1 can also appear in a random-key vector.

@@

Once the decoded solution has been obtained, the fitness value of the solution is computed as in Equation \eqref{eq1}.

\begin{equation}
f_i = z_i + \overbrace{(\mu * n * \text{infeasibility}_i)}^\text{penalty},
\label{eq1}
\end{equation}

@@

In \cite{de2017comparison}, the authors also added a new element to \acs{BRKGA}, an \acs{LS} procedure. This local optimization method is applied to each decoded individual in each generation, but its results are not transferred to the individual, so that diversity is maintained. Thus, the individuals produced by the \acs{LS} are only used to update the best solution found so far if needed. Figure \ref{fig:BRKGA_CC} shows a visual representation of the overall BRKGA optimization process for constrained clustering. For more details on \acs{BRKGA}$_{CC}$ see \cite{de2017comparison}.

\begin{figure}[!h]
	\centering
	\includegraphics[scale=0.31]{gfx/NewProp/SHADE/BRKGA_Diagram.jpg} 
	\caption[BRKGA$_{CC}$ optimization process for constrained clustering.]{\acs{BRKGA}$_{CC}$ optimization process for constrained clustering. \cite{de2017comparison}.}\label{fig:BRKGA_CC}
\end{figure}

\section{The SHADE Optimization Method} \label{sec:SHADE}

\acf{SHADE} is an optimization method based on \acs{DE}. Proposed in \cite{tanabe2013success}, it was an improvement for the JADE model, presented in \cite{zhang2009jade}.

Unlike JADE, which only considers the parameters used in the previous generation, \acs{SHADE} uses a historical record of successful parameters as a mechanism for adapting parameters involved in the process of creating new generations.

@@

\subsection{SHADE Operators}

@@

\subsection{SHADE Parameter Adaptation Method}

@@

	\end{tabular}}
	\caption[Historical memory $M_{CR}$, $M_{F}$ used by SHADE.]{Historical memory $M_{CR}$, $M_{F}$ used by \acs{SHADE} \cite{tanabe2013success}.}
	\label{tab:SHADEmemory}
\end{table}

@@

\subsection{SHADE optimization scheme for constrained clustering} \label{sec:SHADEadapt}

@@

\begin{equation}
f_i = z_i * (\text{infeasibility}_i + 1).
\label{eq14}
\end{equation}

@@

To make these concepts clear we consider a toy dataset and three partitions over it. The dataset from Figure \ref{fig:toydatasets} contains 100 instances and a single ML constraint between two instances from the same class. Partitions $C_0$---displaying the true labels---and $C_1$ do not violate the constraint ($\text{infeasibility} = 0$), whereas $C_2$ does violate it ($\text{infeasibility} = 1$). However, $C_2$ shows a possibility, for later iterations, of adding both instances to their nearest cluster while also satisfying the ML constraint.

\begin{figure}[bth]
	\myfloatalign
	{\includegraphics[width=.41\linewidth]{gfx/NewProp/SHADE/Dataset}} \quad
	%\subfloat[Imagen original]
	{\includegraphics[width=.41\linewidth]{gfx/NewProp/SHADE/C0}} \quad
	%\subfloat[Clustering sin restricciones]
	{\includegraphics[width=.41\linewidth]{gfx/NewProp/SHADE/C1}} \quad
	%\subfloat[Clustering con restricciones]
	{\includegraphics[width=.41\linewidth]{gfx/NewProp/SHADE/C2}} \quad
	\caption[Three partitions over a toy dataset.]{Three partitions over a toy dataset.}
	\label{fig:toydatasets}
\end{figure}

@@

	\caption[Expression and value of fitness functions over three partitions ($\mu = 10$).]{Expression and value of fitness functions over three partitions ($\mu = 10$).}
	\label{tab:fitnessfunctions}
\end{table}

@@

As in \cite{de2017comparison} we apply a local optimization procedure to the individuals of the population, but without transferring its results to the original individual in order to maintain diversity. The difference is that we do not apply it to the whole population but only to a number $P_e$ of individuals considered as the elite of the population, with the aim of preserving a good exploration-exploitation trade-off. The percentage of individuals considered as elite must be determined for each case. Figure \ref{fig:SHADE_CC} shows a representation of the \acs{SHADE}$_{CC}$ optimization process.

\begin{figure}[!h]
	\centering
	\includegraphics[scale=0.31]{gfx/NewProp/SHADE/SHADE_Diagram.jpg} 
	\caption[SHADE$_{CC}$ optimization process for constrained clustering.]{\acs{SHADE}$_{CC}$ optimization process for constrained clustering.}\label{fig:SHADE_CC}
\end{figure}

@@

		}
		\tcc{Whenever $|A| > |P|$, randomly select an individual from $A$ to be deleted so that $|A| \le |P|$}
		\If{$S_{CR} \neq \emptyset$ \textbf{and} $S_{F} \neq \emptyset$}{
			Update $M_{[CR,h]}$ and $M_{[F,h]}$ based on $S_{CR}$ and $S_{F}$\\
			$h \leftarrow (h + 1) \mod |M|$
		}
		$G \leftarrow G + 1$
	}
	\caption[SHADE$_{CC}$]{\acs{SHADE}$_{CC}$}\label{alg:SHADE}
\end{algorithm}

@@

	\caption[SHADE$_{CC}$ Local Search]{\acs{SHADE}$_{CC}$ Local Search}\label{alg:SHADE_LS}
\end{algorithm}

@@

Following the stream of thought introduced at the beginning of this chapter, it is justified to think that metaheuristics constitute a promising approach to fin approximate quality solutions to the tough challenge that constrained clustering constitutes. Metaheuristic algorithms are designed to explore the solution space of a problem with the guidance of a fitness (heuristic) function \cite{Gendreau:2010:HM:1941310}. In this type of approaches the key is to find a good exploration-exploitation tradeoff. This is the reason why the classic \acf{LS} heuristic algorithm derived in the \acf{ILS} metaheuristic algorithm, which introduces periodic perturbations in the exploration of the solution space to escape local optima \cite{lourencco2010iterated}. \acs{ILS}---or variants of it---has been applied to a wide variety of problems, including the traveling salesman \cite{archetti2018iterated}, the quadratic multiple knapsack \cite{avci2017multi} or the vehicle routing problem \cite{chentli2018impact, estrada2019biased}. \acs{ILS} is also very commonly used in the design of hybrid methods with good exploration-exploitation tradeoff, such as its combination with expediting mechanisms \cite{zohali2019reformulation} or with the success history based differential evolution algorithm \cite{zhao2019hybrid}. In this section we propose a new variant of the \acs{ILS} method, the \acs{DILS}, which introduces diversity in the exploration of the solution space in an adaptive and guided way.

\subsection{A Glimpse to Iterated Local Search}

Before defining the \acs{ILS}---from now on \acs{ILS}---,it is necessary to understand the \acs{LS} and the motivation that leads to designing methods with greater exploratory capacity.

An \acs{LS} is a procedure that consists of performing an heuristics-guided iterative improvement. Taking as initial state a well defined solution to the problem, the \acs{LS} uses a neighbor generation operator to explore new solutions. When a solution better than the current one is found, it replaces the current one. In order to evaluate the solutions, the \acs{LS} uses a function that assigns a value to each one of them, this is known as a target function. ***This value will be better the closer the solution that evaluates the optimal solution.***

Both the neighbor generation operator and the target function used by the \acs{LS} must be defined for each problem. The target function $f(\cdot)$ determines how good the solution achieved is with respect to the purpose of the problem. The neighbor generation operator takes as base a solution to the problem $s$, and generates a new solution $s^\prime$ by modifying some of its components in a local way.

The Algorithm \ref{alg:BasicLS} presents a basic scheme for an \acs{LS} procedure. It is worth noting that the condition of line 4 must be adapted if the problem to be solved is one of maximization or minimization. In addition, the general scheme considers that the optimization process ends when the current solution is not improved, however the stop condition can be adapted to each problem.

\begin{algorithm}
	\SetNlSty{textbf}{[}{]}
	\SetNlSkip{0.5em}
	\setstretch{1.2}
	\SetKwFunction{GenerateNeighbor}{GenerateNeighbor}
	\SetKwRepeat{Do}{do}{while}
	\KwIn{Initial solution $s$}
	\BlankLine
	\While{$improvement$}{
		$improvement \leftarrow$ \texttt{false} \\

		$s^\prime \leftarrow $ \GenerateNeighbor{$s$}

		\If{$f(s^\prime) < f(s)$}{

			$s \leftarrow s^\prime$\\
			$improvement \leftarrow$ \texttt{true} \\

		}	
	}
	\BlankLine
	\KwRet ($s$)

@@

With the \acs{LS} defined in this way, it is easy to see that the solution achieved will be the optimal local $s^*$ closest to the initial solution $s$. Figure \ref{img:LS} shows a pictorical representation of the above.

\begin{figure}[!h]
	\centering
	\includegraphics[scale=0.25]{gfx/NewProp/DILS/LS.png}
	\caption[Local optimum achieved by an LS procedure]{Local optimum achieved by an \acs{LS} procedure}\label{img:LS}
\end{figure}

It is therefore necessary to appeal to techniques that introduce diversity in the search to avoid the problem of local optimal. This is exactly what \acs{ILS} is trying to do.

With \acs{ILS}, once we have reached a local optimum $s^*$, we apply a perturbation to it that, with a high probability, will result in a new solution $s^\prime$ that will not be a local optimum and will be far enough away from $s^*$. After that, we will apply an \acs{LS} procedure to find a new local optimal $s^{*\prime}$. Figure \ref{img:ILS} shows a representation of the above concepts.

\begin{figure}[!h]
	\centering
	\includegraphics[scale=0.25]{gfx/NewProp/DILS/ILS.png}
	\caption[ILS perturbation over LS local optima]{\acs{ILS} perturbation over \acs{LS} local optima}\label{img:ILS}
\end{figure}

It should be noted that, if the disturbance applied to $s*$ is too small, the most likely is that the local optimum closest to $s^\prime$ is $s^*$ and therefore the solution space will not be explored effectively. On the other hand, if the disturbance is too big, there will be no bias in the generation of new solutions, and therefore the procedure becomes a random reboot-type method.

The method for generating the perturbation applied to the local optimum $s^*$ may vary depending on the problem, and may be based on previous solutions. In this way, it is possible to keep a history of the solutions reached so far, so they can be used when applying the perturbation. Algorithm \ref{alg:ILS} summarizes the \acs{ILS} overall optimization process.

\begin{algorithm}
	\SetNlSty{textbf}{[}{]}
	\SetNlSkip{0.5em}
	\setstretch{1.2}
	\SetKwFunction{LocalSearch}{LocalSearch}
	\SetKwFunction{Perturbation}{Perturbation}
	\SetKwRepeat{Do}{do}{while}
	\KwIn{Initial solution $s$}
	\BlankLine
	$s^* \leftarrow$ \LocalSearch{$s$}\\
	\While{Termination criteria are not met}{

		$s^\prime \leftarrow $ \Perturbation{$s^*$, $history$}\\
		$s^{*\prime} \leftarrow$ \LocalSearch{$s^\prime$}

		\If{$f(s^{*\prime}) < f(s^*)$}{

			$s^* \leftarrow s^{*\prime}$

		}	
	}
	\BlankLine
	\KwRet ($s*$)

@@

In the \acs{ILS} scheme described in Algorithm \ref{alg:ILS} the acceptance criterion only use the difference in the cost of the solutions to be compared to choose between them. This is the most simple scheme and does not depends on previous solutions recorded in the history.

@@

The \acf{DILS} is a new variant of the classic \acs{ILS} method. Its goal is to perform a search in the solution space to find the solution with the best fitness value (given by fitness function $f$) by introducing diversity in an adaptive and guided way. While \acs{ILS} works with a single individual, \acs{DILS} keeps two of them $\{m_b, m_w\}$ in memory at all times, which allows it to guide diversity-inducing methods and to avoid local optima. The individual $m_b$ is the one providing the best fitness value, whereas $m_w$ provides the worst fitness value at the end of each stage (generation) of the optimization process.

In each generation $G$ of the optimization process \acs{DILS} builds a new individual by applying a recombination operator to $m_b$ and $m_w$. After that, it applies a strong mutation operator to the newly generated individual. This is the individual \acs{DILS} applies the \acs{LS} procedure to in order to improve its fitness value, resulting in the trial individual $m_t$. The trial individual $m_t$ replaces the worst of its predecessors, $m_w$, if it achieves a better fitness value; this operation represents the acceptance criterion for \acs{DILS}. Equation \ref{eq2.1} displays the expression for this criterion in the case of a minimization problem.

\begin{equation}
m_{w,G} = \left\{ \begin{array}{lc}
m_t &   \text{if} \;\; f(m_t) < f(m_w)\\
m_{w,G} &  \text{otherwise}
\end{array}
\right..
\label{eq2.1}
\end{equation}

>>>>

Finding the optimal partition in a dataset according to any kind of reasonable criteria is known to be a $\mathbf{NP}$-hard problem. As we have already mentioned in Section \ref{sec:FeasibilityProblem}, the use of constraints may modify the complexity of the clustering problem, making it $\mathbf{NP}$-complete and therefore intractable \cite{davidson2005clustering}. Bearing this in mind, in this chapter we propose two new approaches to find approximate solutions to the constrained clustering problem. In Section \ref{sec:CCSHADE} a \acf{DE} based approach is presented, while in Section \ref{sec:DILS} a new \acf{LS} variant is discussed, as well as its application to constrained clustering.

@@

\acs{DE} is an evolution-based algorithm that has proven to be excellent in real-domain problem solving \cite{das2011differential}. In this section we propose the first \acs{DE} application to find quality solutions for the constrained clustering problem. In particular, we will take as basis for our proposal one of the \acs{DE} variants which showed excellent behavior in worldwide competitions, known as \acs{SHADE} \cite{molina2018insight}. We will make use of the Random-key concept from \acs{BRKGA}, along with proposing a new fitness function, to build the already mentioned \acs{DE} application.

\subsection{A Glimpse into Differential Evolution}

@@

To guide the optimization process, we use a fitness function $f(\cdot)$. The task of \acs{DE} is to seek the parameter vector $p_i^*$ that minimizes a function such that $f(p_i)(f: \Omega \subseteq \mathfrak{R}^{v} \rightarrow \mathfrak{R})$, where $f(p_i^*) < f(p_i)$ for all $p_i \in \Omega$ with $\Omega$ a nonempty finite set that is the search domain.

\begin{figure}[!h]
	\centering
	\includegraphics[scale=0.3]{gfx/NewProp/SHADE/DEop.png}
	\caption[The optimization process of \acs{DE}.]{The optimization process of \acs{DE} \cite{das2011differential}.}\label{fig:DE}
\end{figure}

@@

Even though \acs{DE} was a major breakthrough in terms of real optimization, many variants have been developed since its invention that provide better results. We will focus on the \acs{SHADE} variant, which adaptively optimizes some of the parameters of \acs{DE}. To carry out this task \acs{SHADE} uses a historic record of values for these parameters that allows it to set them in favor of the optimization process, based on how good the algorithm behaved with certain parameter values in the past (see Section \ref{sec:SHADE}).

\subsection[Brief Review of the \acsfont{BRKGA} Algorithm]{Brief Review of the BRKGA Algorithm} \label{sec:brkga}

@@

In a \acs{GA}, a population $P$ of a given number of solutions \sloppy $p_i = \{p_{[i,1]}, \cdots, p_{[i,v]}\}$ is first initialized; these solutions are also called chromosomes or individuals. Each element of a chromosome is called gene and the value it takes is known as allele. To simulate the natural processes of evolution and natural selection, the set of evolutionary operators that \acsp{GA} make use of is similar to that used by \acs{DE}: crossover, mutation and selection. The crossover operator consists in combining two different chromosomes (parents) from the population and obtaining a new chromosome (offspring) which inherits alleles for its genes from both parents. The mutation operator is used by \acsp{GA} to avoid local optima, and it consists in introducing random changes into the population to favor diversity. The selection operator is used to determine whether a chromosome is suitable to survive for the next generation, and to accomplish this task it employs a fitness function $f(\cdot)$ that assigns a value to each chromosome to assess its quality as a solution for the problem. \acsp{GA} apply these operators through a series of generations to hopefully obtain a solution for the problem to solve that is close to the optimal \cite{fausto2019ants}.

\acf{RKGA} emerged in \cite{bean1994genetic} as a solid approach to highly constrained problems. It uses vectors of random numbers to represent solutions to the problem (individuals) which are used as keys to obtain the actual solutions. Random-key vectors are sampled from the space $[0,1]^{v}$, which is used as a surrogate space for the literal solution space. Points in the random-key space are transferred to the literal solution space via a decoder for fitness evaluation. A decoder is a deterministic procedure that allows us to map random-key vectors to the literal solution space. We can build a decoder that always provides as a result a feasible solution to the problem we want to solve, thus avoiding the feasibility problem that we found in classic \acsp{GA}.

The \acf{BRKGA} was first proposed in \cite{gonccalves2011biased} as a variant of the \acs{RKGA}. In the \acs{BRKGA} optimization process, first, the population $P$ of $|P|$ vectors of random-keys is initialized; this will be the initial generation. Each random-key vector $p_i$ has $v$ random-keys in it. The population is sorted by the fitness value $f(p_i)$ of each individual to select the first $P_e$ individuals, this is, the elite of the population. The elite will be preserved in the next generation without modification. To introduce diversity, a number $P_m$ of new random-keys vectors are also included in the next generation. The remaining individuals of the new generation ($|P| - P_e - P_m$) are obtained through crossovers between elite and non-elite parents.

@@

\subsection[\acsfont{BRKGA} Optimization Scheme for Constrained Clustering]{BRKGA Optimization Scheme for Constrained Clustering} \label{sec:AdaptationofBRKGA}

@@

In this scheme each random-key vector has $n$ random-keys in it ($v = n$). The decoder divides the interval $[0,1]$ in $k$ intervals, so there exists a correspondence between each random-key (instance $x_i$) and the integer (label $l_i$) corresponding to the interval which it lies in. Table \ref{tab:decodingrk} shows an example of random-key decoding for a dataset withteninstances ($n$ = 10) and $k = 3$. Note that extreme values 0 and 1 can also appear in a random-key vector.

@@

Once the decoded solution has been obtained, the fitness value of the solution (individual $p_i$) is computed as in Equation \eqref{eq1}.

\begin{equation}
f_1(p_i) = z_i + \overbrace{(\mu * n * \text{infeasibility}_i)}^\text{penalty},
\label{eq1}
\end{equation}

@@

In \cite{de2017comparison}, the authors also added a new element to \acs{BRKGA}, an \acs{LS} procedure. This local optimization method is applied to each decoded individual in each generation, but its results are not transferred to the individual, so that diversity is maintained. Thus, the individuals produced by the \acs{LS} are only used to update the best solution found so far if needed. Figure \ref{fig:BRKGA_CC} shows a visual representation of the overall \acs{BRKGA} optimization process for constrained clustering. For more details on \acs{BRKGA}$_{CC}$ see \cite{de2017comparison}. Arrows in red indicate population sorting by fitness.

\begin{figure}[!h]
	\centering
	\includegraphics[scale=0.31]{gfx/NewProp/SHADE/BRKGA_Diagram.jpg} 
	\caption[\acs{BRKGA}$_{CC}$ optimization process for constrained clustering.]{\acs{BRKGA}$_{CC}$ optimization process for constrained clustering \cite{de2017comparison}.}\label{fig:BRKGA_CC}
\end{figure}

\subsection[The \acsfont{SHADE} Optimization Method]{The SHADE Optimization Method} \label{sec:SHADE}

\acf{SHADE} is an optimization method based on \acs{DE}. Proposed in \cite{tanabe2013success}, it was an improvement for the \acsfont{JADE} model, presented in \cite{zhang2009jade}.

Unlike \acsfont{JADE}, which only considers the parameters used in the previous generation, \acs{SHADE} uses a historical record of successful parameters as a mechanism for adapting parameters involved in the process of creating new generations.

@@

\subsubsection[\acsfont{SHADE} Operators]{SHADE Operators}

@@

\subsubsection[\acsfont{SHADE} Parameter Adaptation Method]{SHADE Parameter Adaptation Method}

@@

	\end{tabular}}
	\caption[Historical memory $M_{CR}$, $M_{F}$ used by \acs{SHADE}.]{Historical memory $M_{CR}$, $M_{F}$ used by \acs{SHADE} \cite{tanabe2013success}.}
	\label{tab:SHADEmemory}
\end{table}

@@

\subsection[\acsfont{SHADE} Optimization Scheme for Constrained Clustering]{SHADE Optimization Scheme for Constrained Clustering} \label{sec:SHADEadapt}

@@

\begin{equation}
f_2(p_i) = z_i * (\text{infeasibility}_i + 1).
\label{eq14}
\end{equation}

@@

To make these concepts clear we consider a toy dataset and three partitions over it. The dataset from Figure \ref{fig:toydatasets} contains 100 instances and a single \acs{ML} constraint between two instances from the same class. Partitions $C_0$---displaying the true labels---and $C_1$ do not violate the constraint ($\text{infeasibility} = 0$), whereas $C_2$ does violate it ($\text{infeasibility} = 1$). However, $C_2$ shows a possibility, for later iterations, of adding both instances to their nearest cluster while also satisfying the \acs{ML} constraint.

\begin{figure}[bth]
	\myfloatalign
	{\includegraphics[width=.41\linewidth]{gfx/NewProp/SHADE/Dataset}} \quad
	%\subfloat[Imagen original]
	{\includegraphics[width=.41\linewidth]{gfx/NewProp/SHADE/C0}} \quad
	%\subfloat[Clustering sin restricciones]
	{\includegraphics[width=.41\linewidth]{gfx/NewProp/SHADE/C1}} \quad
	%\subfloat[Clustering con restricciones]
	{\includegraphics[width=.41\linewidth]{gfx/NewProp/SHADE/C2}} \quad
	\caption{Three partitions over a toy dataset.}
	\label{fig:toydatasets}
\end{figure}

@@

	\caption{Expression and value of fitness functions over three partitions ($\mu = 10$).}
	\label{tab:fitnessfunctions}
\end{table}

@@

As in \cite{de2017comparison} we apply a local optimization procedure to the individuals of the population, but without transferring its results to the original individual in order to maintain diversity. The difference is that we do not apply it to the whole population but only to a number $P_e$ of individuals considered as the elite of the population, with the aim of preserving a good exploration-exploitation trade-off. The percentage of individuals considered as elite must be determined for each case. Figure \ref{fig:SHADE_CC} shows a representation of the \acs{SHADE}$_{CC}$ optimization process. Arrows in red indicate population sorting by fitness.

\begin{figure}[!h]
	\centering
	\includegraphics[scale=0.3]{gfx/NewProp/SHADE/SHADE_Diagram.jpg} 
	\caption{\acs{SHADE}$_{CC}$ optimization process for constrained clustering.}\label{fig:SHADE_CC}
\end{figure}

@@

		}
		\tcc{Whenever $|A| > |P|$, randomly select an individual from $A$ to be deleted so that $|A| \le |P|$}
		\If{$S_{CR} \neq \emptyset$ \textbf{and} $S_{F} \neq \emptyset$}{
			Update $M_{[CR,h]}$ and $M_{[F,h]}$ based on $S_{CR}$ and $S_{F}$\\
			$h \leftarrow (h + 1) \mod |M|$
		}
		$G \leftarrow G + 1$
	}
	\caption{\acs{SHADE}$_{CC}$}\label{alg:SHADE}
\end{algorithm}

@@

	\caption{\acs{SHADE}$_{CC}$ Local Search}\label{alg:SHADE_LS}
\end{algorithm}

@@

Following the stream of thought introduced at the beginning of this chapter, it is justified to think that metaheuristics constitute a promising approach to find approximate quality solutions to the tough challenge that constrained clustering represents. Metaheuristic algorithms are designed to explore the solution space of a problem with the guidance of a fitness (heuristic) function \cite{Gendreau:2010:HM:1941310}. In this type of approaches the key is to find a good exploration-exploitation tradeoff. This is the reason why the classic \acf{LS} heuristic algorithm derived in the \acf{ILS} metaheuristic algorithm, which introduces periodic perturbations in the exploration of the solution space to escape local optima \cite{lourencco2010iterated}. \acs{ILS} (or variants of it) has been applied to a wide variety of problems, including the traveling salesman \cite{archetti2018iterated}, the quadratic multiple knapsack \cite{avci2017multi} or the vehicle routing problem \cite{chentli2018impact, estrada2019biased}. \acs{ILS} is also very commonly used in the design of hybrid methods with good exploration-exploitation tradeoff, such as its combination with expediting mechanisms \cite{zohali2019reformulation} or with the success history based differential evolution algorithm \cite{zhao2019hybrid}. In this section we propose a new variant of the \acs{ILS} method, the \acs{DILS}, which introduces diversity in the exploration of the solution space in an adaptive and guided way.

\subsection{A Glimpse into Iterated Local Search}

Before defining \acs{ILS}, it is necessary to understand \acs{LS} and the motivation behind the design of methods with greater exploratory capabilities.

\acs{LS} is a procedure that consists in performing an heuristics-guided iterative improvement. With a valid solution to the problem as an initial state, \acs{LS} uses a neighbor generation operator to explore new solutions. The current solution is replaced when a better one is found. In order to evaluate solutions, a function to assess their suitability, known as \textit{fitness function}, is used. Ideally, the closer a solution is to the global optimum, the better its fitness value will be.

Both the neighbor generation operator and the fitness function used by \acs{LS} must be defined for each particular problem. The fitness function $f(\cdot)$ determines how good the solution achieved is with respect to the purpose of the problem. The neighbor generation operator attempts to improve upon a solution $S$ to the problem by modifying a few of its components, and thus generating a new solution $S^\prime$.

The Algorithm \ref{alg:BasicLS} presents a basic scheme for an \acs{LS} procedure. It is worth noting that the condition of line 4 must be adapted depending on whether the problem to be solved is one of maximization or minimization. In addition, the general scheme assumes that the optimization process ends when the current solution is not improved. However, the stop condition can be adapted to each problem.

\begin{algorithm}
	\SetNlSty{textbf}{[}{]}
	\SetNlSkip{0.5em}
	\setstretch{1.2}
	\SetKwFunction{GenerateNeighbor}{GenerateNeighbor}
	\SetKwRepeat{Do}{do}{while}
	\KwIn{Initial solution $S$}
	\BlankLine
	\While{$improvement$}{
		$improvement \leftarrow$ \texttt{false} \\

		$S^\prime \leftarrow $ \GenerateNeighbor{$S$}

		\If{$f(S^\prime) < f(S)$}{

			$S \leftarrow S^\prime$\\
			$improvement \leftarrow$ \texttt{true} \\

		}	
	}
	\BlankLine
	\KwRet ($S$)

@@

With \acs{LS} defined in this way, it is easy to see that the solution achieved will be the local optimum $S^*$ that is closest to the initial solution $S$. Figure \ref{img:LS} shows a pictorial representation of the above.

\begin{figure}[!h]
	\centering
	\includegraphics[scale=0.25]{gfx/NewProp/DILS/LS.png}
	\caption{Local optimum achieved by an \acs{LS} procedure.}\label{img:LS}
\end{figure}

It is therefore necessary to turn to techniques that introduce diversity in the search procedure to avoid the problem of local optimum. This is exactly what \acs{ILS} is trying to do. With \acs{ILS}, once we have reached a local optimum $S^*$, we apply a perturbation to it that will result in a new solution $S^\prime$. With a high probability, $S^\prime$ will not be another local optimum and instead will be far enough away from $S^*$. After that, we will apply an \acs{LS} procedure to find a new local optimum $S^{*\prime}$. Figure \ref{img:ILS} shows a representation of the aforementioned concepts.

\begin{figure}[!h]
	\centering
	\includegraphics[scale=0.25]{gfx/NewProp/DILS/ILS.png}
	\caption{\acs{ILS} perturbation over an \acs{LS} local optimum.}\label{img:ILS}
\end{figure}

It should be noted that, if the perturbation applied to $S*$ is too small, most likely the local optimum closest to $S^\prime$ is again $S^*$ and therefore the exploration of the solution space is not be effective. On the other hand, if the perturbation is too big, there will be no bias in the generation of new solutions, and therefore the procedure becomes a random-reboot method.

The generation of perturbations for the local optimum $S^*$ may vary depending on the problem, and may be based on previous solutions. In this way, it is possible to keep a history of solutions reached so far, so they can be used when applying the perturbation. Algorithm \ref{alg:ILS} summarizes the overall optimization process of \acs{ILS}.

\begin{algorithm}
	\SetNlSty{textbf}{[}{]}
	\SetNlSkip{0.5em}
	\setstretch{1.2}
	\SetKwFunction{LocalSearch}{LocalSearch}
	\SetKwFunction{Perturbation}{Perturbation}
	\SetKwRepeat{Do}{do}{while}
	\KwIn{Initial solution $S$}
	\BlankLine
	$S^* \leftarrow$ \LocalSearch{$S$}\\
	\While{Termination criteria are not met}{

		$S^\prime \leftarrow $ \Perturbation{$S^*$, $history$}\\
		$S^{*\prime} \leftarrow$ \LocalSearch{$S^\prime$}

		\If{$f(S^{*\prime}) < f(S^*)$}{

			$S^* \leftarrow S^{*\prime}$

		}	
	}
	\BlankLine
	\KwRet ($S*$)

@@

In the \acs{ILS} scheme described in Algorithm \ref{alg:ILS} the acceptance criterion only uses the difference in the cost of the compared solutions to choose between them. This is the simplest scheme and does not rely on the history of previous solutions.

@@

The Dual Iterative Local Search (DILS) is a new variant of the classic ILS method. Its goal is to perform a search in the solution space to find the solution with the best fitness value (given by fitness function $f$) by introducing diversity in an adaptive and guided way. While ILS works with a single individual, DILS keeps two of them $\{m_b, m_w\}$ in memory at all times, which allows it to guide diversity-inducing methods and to avoid local optima. The individual $m_b$ is the one providing the best fitness value, whereas $m_w$ provides the worst fitness value at the end of each stage (generation) of the optimization process.

In each generation $G$ of the optimization process DILS builds a new individual by applying a recombination operator to $m_b$ and $m_w$. After that, it applies a strong mutation operator to the newly generated individual. This is the individual DILS applies the LS procedure to in order to improve its fitness value, resulting in the trial individual $m_t$. The trial individual $m_t$ replaces the worst of its predecessors, $m_w$, if it achieves a better fitness value; this operation represents the acceptance criterion for DILS. Equation \ref{eq1.2} displays the expression for this criterion in the case of a minimization problem.

\begin{equation}
m_{w,G} = \left\{ \begin{array}{lc}
m_t &   \text{if} \;\; f(m_t) < f(m_w)\\
m_{w,G} &  \text{otherwise}
\end{array}
\right..
\label{eq1.2}
\end{equation}

Now, the process described so far will most likely fall into a local optimum, effectively stopping the exploration of the solution space. To avoid it DILS implements a reinitialization method for $m_w$ based on the difference between the two individuals it keeps in memory ($\{m_b, m_w\}$). Equation \ref{eq2.2} shows the reinitialization criterion for a minimization problem. RandInit$()$ is a function that returns a randomly initialized individual and $\xi \in [-1,1]$ is the parameter that controls the tolerance of the reinitialization method.
